---
title: "Diabetes Prediction"
output: html_document
---

DSCI 310, group 10: Niki Duan, Vivian Pu, Nathan Sihombing, Yun Tian

# Abstract Summary

This study investigates the use of machine learning techniques to predict diabetes status using clinical and demographic data from the Pima Indian Diabetes dataset. The dataset includes physiological variables such as glucose level, body mass index (BMI), age, and number of pregnancies.

A random forest classification model was developed to capture nonlinear relationships among predictors and improve prediction accuracy. The model demonstrated strong performance on the test dataset, achieving an accuracy of 88.9%, precision of 89.5%, recall of 94.0%, and an F1 score of 91.7%. The high recall indicates that the model is particularly effective at identifying individuals with diabetes, making it well-suited for screening and early detection applications.

Despite these promising results, the study has limitations. The dataset includes only female participants, which may restrict the generalizability of the model to broader populations. Overall, this analysis demonstrates that machine learning models, particularly ensemble methods like random forests, can effectively support diabetes risk prediction, though further validation across diverse populations is necessary before clinical implementation.

# Introduction
## Background Information
Diabetes is a chronic metabolic disease that affects millions of people worldwide, with significant health consequences if not properly managed (Spinelli et al., 2020). Early detection of diabetes is critical for preventing complications such as cardiovascular disease, kidney dysfunction, nerve damage, and vision problems (Spinelli et al., 2020). Women may face unique challenges related to diabetes, including an increased risk of pregnancy-related complications and a higher susceptibility to certain cancers (Berg, 2005).

Accurate prediction of diabetes using clinical and demographic data can help identify individuals at risk, enabling timely interventions and better management of the disease.

In this project, we aim to use statistical and machine learning methods to predict the presence of diabetes in individuals based on relevant clinical measurements and features from the dataset.

## Research Question
The question we will try to answer with this project is: "Can we predict the presence of diabetes in individuals based on clinical and demographic features?"

This question sets up the possibility of conducting statistical analyses and building predictive models to determine which features are most strongly associated with diabetes and to estimate the probability of an individual having the disease.

We have framed our question based on previous studies that have shown certain clinical indicators, such as high blood glucose, BMI, age, and family history, are strongly associated with diabetes risk. For example, the Pima Indian study dataset has been widely used to develop predictive models for diabetes, highlighting the role of these risk factors in disease onset (Chang et al., 2022, Roumie et al., 2020).


## Dataset Used
The unprocessed dataset was acquired from the UCI Machine Learning Repository and is originally from the National Institute of Diabetes and Digestive and Kidney Diseases. The version used in this project was processed by a contributor on OpenML (https://www.openml.org/search?type=data&status=active&id=43483) and contains information on blood pressure, diabetes status, and other health-related variables of females. Based on the diagnostic measurements included in the dataset, the objective is to predict whether a patient has diabetes. The dataset was selected from a larger database using several constraints. Specifically, all patients are female, 21 years of age or older, and of Pima Indian heritage.

The dataset attributes are as follows:

- Pregnancies: Number of pregnancies
- Glucose: Blood glucose level
- BloodPressure: Blood pressure measurement
- SkinThickness: Skinfold thickness
- Insulin: Blood insulin level
- BMI: Body Mass Index
- DiabetesPedigreeFunction: Diabetes pedigree function (a score indicating genetic predisposition)
- Age: Age of the patient
- Outcome: Final result (1 = diabetes, 0 = no diabetes)

# EDA


```{r}
# loading all the required libraries. 
library(tidyverse)
library(dplyr)
library(tidymodels) 
library(gridExtra)
library(repr)
library(GGally)
options(repr.matrix.max.rows = 6)
set.seed(1)
```

```{r}
library(OpenML)
# OpenML dataset ID
dataset_id <- 43483

# Download
oml_data <- getOMLDataSet(data.id = dataset_id)

# Extract as data.frame
df <- oml_data$data

head(df)
```
**Table 1: `df` the entire dataset, having 9 variables.**

The df is already very clean with no missing values as shown below.

```{r}
colSums(is.na(df))

# Columns to check if they contain 0 (which indicates missing values)
cols <- c("Glucose","BloodPressure","SkinThickness","Insulin","BMI")

# Check which columns have 0
sapply(df[ , cols], function(x) any(x == 0))
```

```{r}
summary(df)
```
**Table 2: Baseline characteristics of participants included in the diabetes dataset, presented as minimum, 1st quartile, median, mean, 3rd quartile, and maximum values for each variable**

DDataset Overview:
- Total observations: 768  
- Target variable: Outcome (0 = no diabetes, 1 = diabetes)  
- Proportion of patients with diabetes: 34.9%

About 34.9% of the patients in the dataset are labeled as having diabetes.

A potential challenge in this dataset is class imbalance. Approximately 35% of observations are labeled as diabetic, while the remaining 65% are labeled as non-diabetic. Because of this imbalance, accuracy alone is not enough to assess model performance. For that reason, we also report precision, recall, and F1 score, which provide a better picture of how well the model identifies the minority diabetic class.

```{r, fig.width=12, fig.height=10}
# Ensure Outcome is a factor for coloring
df$Outcome <- factor(df$Outcome, levels=c(0,1), labels=c("No","Yes"))

# Select numeric columns and the target
plot_df <- df %>% select(Pregnancies, Glucose, BloodPressure,
                         SkinThickness, Insulin, BMI,
                         DiabetesPedigreeFunction, Age, Outcome)

# Generate pair plot with coloring by Outcome
ggpairs(plot_df, 
        aes(color = Outcome, alpha = 0.5),
        columns = 1:8,   # numeric columns
        upper = list(continuous = wrap("cor", size = 3)), # show correlation in upper panel
        lower = list(continuous = "points"),
        diag = list(continuous = "densityDiag")) +
  theme_minimal() +
  ggtitle("Pairwise Relationships Between Features and Outcome")
```
**Figure 1: Matrix of scatter plots (lower panel), density plots (diagonal), and correlation coefficients (upper panel) for the eight predictor variables in the Pima Indian Diabetes dataset. Points are color-coded by diabetes status (blue = No diabetes, red = Diabetes), with overlaid density curves showing the distribution of each variable for the two outcome groups.**

The pair plot reveals that glucose concentration has the strongest association with diabetes outcome (r = 0.53, p < 0.001), followed by BMI (r = 0.29, p < 0.001), age (r = 0.27, p < 0.001), and number of pregnancies (r = 0.22, p < 0.001), while other variables such as insulin, skin thickness, blood pressure, and diabetes pedigree function show weaker but still statistically significant correlations. Notable inter-feature relationships include strong positive correlations between age and pregnancies (r = 0.54) and between BMI and skin thickness (r = 0.54), as well as moderate correlations among insulin, glucose, and BMI, indicating expected physiological links. The density plots along the diagonal further show that individuals with diabetes tend to have higher glucose levels, greater BMI, and are older compared to those without diabetes.

# Methods & Results


To predict diabetes status, we performed a supervised classification analysis using a random forest model in R (R Core Team, 2019). And the following R packages were used to perform the analysis: knitr (Xie 2014), tidyverse (Wickham 2017), and Quarto (Allaire et al 2022).

First, the dataset was imported into R as a data frame. The outcome variable representing diabetes status was converted to a factor to ensure that the task was treated as a classification problem rather than regression. All predictor variables were retained as features for model training.

The dataset was then partitioned into training and testing subsets using an 80/20 split. Stratified sampling was applied to preserve the class distribution of the diabetes outcome variable in both subsets. This ensured that the model was trained and evaluated on representative data.

A random forest classifier was trained using the training dataset. The model was fit using 500 decision trees, and the number of variables randomly sampled at each split was set to the default value (the square root of the number of predictors). A decision tree is a supervised learning model that makes predictions by recursively splitting the data into subsets based on feature values, forming a tree-like structure of decision rules that lead to a final outcome. Random forest was selected because it is robust to nonlinear relationships, automatically captures interactions between variables, and is less prone to overfitting compared to a single decision tree.

After training, the fitted model was used to generate predictions on the held-out test dataset. These predictions were compared against the true diabetes labels to evaluate model performance.

Model performance was assessed using a confusion matrix. From the confusion matrix, several evaluation metrics were calculated, including:

- Accuracy, representing the proportion of correctly classified observations.
- Precision, measuring the proportion of predicted positive cases that were truly positive.
- Recall (sensitivity), measuring the proportion of actual positive cases correctly identified.
- F1 score, representing the harmonic mean of precision and recall.

The confusion matrix was visualized as a heatmap to provide a graphical summary of classification performance. Additionally, a summary table was created to report the accuracy, precision, recall, and F1 score of the model.

All analyses were conducted in R using the randomForest and caret packages.

```{r}
# Load required libraries
library(randomForest)
library(caret)

df$Outcome <- as.factor(df$Outcome)

# split training and testing set
train_index <- createDataPartition(df$Outcome, p = 0.8, list = FALSE)
train_data <- df[train_index, ]
test_data  <- df[-train_index, ]

rf_model <- randomForest(
  Outcome ~ ., 
  data = train_data,
  ntree = 500,
  importance = TRUE
)

# apply the model on the test set
predictions <- predict(rf_model, newdata = test_data)

cm <- confusionMatrix(predictions, test_data$Outcome)
cm_table <- as.data.frame(cm$table)

colnames(cm_table) <- c("Prediction", "Reference", "Freq")

# plot the confusion matrix
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 6) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Confusion Matrix - Random Forest",
    x = "Actual",
    y = "Predicted"
  ) +
  theme_minimal(base_size = 14)
```

**Figure 2. Confusion matrix for the random forest model predicting diabetes status on the test dataset. The matrix shows the number of true positives, true negatives, false positives, and false negatives, comparing predicted classifications to the actual outcomes.**

```{r}
# Overall accuracy
accuracy <- cm$overall["Accuracy"]

# For binary classification (Positive class is usually the first level)
precision <- cm$byClass["Precision"]
recall    <- cm$byClass["Recall"]
f1        <- cm$byClass["F1"]

metrics_table <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F1 Score"),
  Value  = c(accuracy, precision, recall, f1)
)

metrics_table
```

**Table 3. Performance metrics for the random forest model predicting diabetes on the test dataset, including accuracy, precision, recall, and F1 score.**

The model performed well on the held-out test set, with an accuracy of 88.9%, precision of 89.5%, recall of 94.0%, and an F1 score of 91.7%. The high recall suggests that the model identified most diabetes cases in this particular test split, which is valuable in a screening-oriented setting where missed positive cases are especially concerning. The precision and F1 score also suggest a reasonably strong balance between identifying true diabetes cases and limiting false positives. However, these results should be interpreted as performance on this dataset and test split, rather than as evidence that the model would perform equally well in broader clinical settings.
# Discussion

## Interpretation of Findings

In this analysis, we developed a random forest classification model to predict diabetes status using clinical and physiological predictors. On the held-out test set, the model achieved an accuracy of 88.9%, precision of 89.5%, recall of 94.0%, and an F1 score of 91.7%. These results suggest that the model performed well on this dataset, especially in terms of recall, meaning that it identified most of the diabetes cases in the test set.

Random forest models are well-suited for structured clinical data because they can capture nonlinear relationships and interactions between variables without strong assumptions. The findings are consistent with the exploratory analysis, where glucose, BMI, age, and pregnancies appeared to be among the most informative predictors.

## Limitations

This analysis has several limitations. First, the dataset includes only female participants, which limits how well the model may generalize to males or broader populations.

Second, the model was evaluated using a single 80/20 train-test split. This means the reported performance depends on one specific partition of the data and may vary with different splits.

Third, the dataset shows class imbalance, with fewer diabetes cases than non-diabetes cases. While we reported precision, recall, and F1 score to address this, we did not apply additional techniques such as resampling or class weighting.

Finally, the model was not validated on external datasets, so its performance in real-world settings remains uncertain.

## Future Work

Future work could improve this analysis by using cross-validation to obtain more stable performance estimates. Comparing the random forest model with other models such as logistic regression or gradient boosting could also provide additional insights.

Addressing class imbalance through resampling or threshold tuning may further improve model performance. In addition, testing the model on external datasets would help evaluate generalizability.

## Conclusion

Overall, this study shows that machine learning models can predict diabetes status reasonably well within this dataset. The model achieved strong performance, particularly in recall, suggesting potential value for identifying individuals at elevated risk. However, due to dataset limitations, class imbalance, the use of a single train-test split, and lack of external validation, the results should be interpreted cautiously.

# References 
- “OpenML.” Openml.org, 2025, www.openml.org/search?type=data&id=43483&sort=runs&status=active.
- Chang, V., Bailey, J., Xu, Q. A., & Sun, Z. (2022). Pima Indians diabetes mellitus classification based on machine learning (ML) algorithms. Neural computing & applications, 1–17. Advance online publication. https://doi.org/10.1007/s00521-022-07049-z
- Roumie, C. L., Hung, A. M., Russell, G. B., Basile, J., Kreider, K. E., Nord, J., Ramsey, T. M., Rastogi, A., Sweeney, M. E., Tamariz, L., Kostis, W. J., Williams, J. S., Zias, A., Cushman, W. C., & SPRINT Research Group (2020). Blood Pressure Control and the Association With Diabetes Mellitus Incidence: Results From SPRINT Randomized Trial. Hypertension (Dallas, Tex. : 1979), 75(2), 331–338. https://doi.org/10.1161/HYPERTENSIONAHA.118.12572
- Kautzky-Willer, A., Leutner, M., & Harreiter, J. (2023). Sex differences in type 2 diabetes. Diabetologia, 66(6), 986–1002. https://doi.org/10.1007/s00125-023-05891-x
- Spinelli, R., Parrillo, L., Longo, M. et al. Molecular basis of ageing in chronic metabolic diseases. J Endocrinol Invest 43, 1373–1389 (2020). https://doi.org/10.1007/s40618-020-01255-z
- Berg M. (2005). Pregnancy and diabetes: how women handle the challenges. The Journal of perinatal education, 14(3), 23–32. https://doi.org/10.1624/105812405X57552
- Allaire, J. J., Charles Teague, Carlos Scheidegger, Yihui Xie, and Christophe Dervieux. 2022. “Quarto.” https://doi.org/10.5281/zenodo.5960048.
- R Core Team. 2019. R: A Language and Environment for Statistical Computing. Vienna, Austria: R Foundation for Statistical Computing. https://www.R-project.org/.
- Wickham, Hadley. 2017. Tidyverse: Easily Install and Load the ’Tidyverse’. https://CRAN.R-project.org/package=tidyverse.
- Xie, Yihui. 2014. “Knitr: A Comprehensive Tool for Reproducible Research in R.” In Implementing Reproducible Computational Research, edited by Victoria Stodden, Friedrich Leisch, and Roger D. Peng. Chapman; Hall/CRC. http://www.crcpress.com/product/isbn/9781466561595.